Community-Centric Innovation

### 2014-03-01 paper.tex
added pdf

Community wireless are particularly useful in underserved areas with little presence of commercial operators.
In rural, distant or poor areas community networks can be the only viable alternative to access the Internet.
In these places, the inhabitants have a strong incentive to spend time and energies in acquiring the necessary skills required to deploy and operate a network.

@@

There are a large number of community networks and they are very heterogeneous.
The use different technologies such as wireless point-to-point links, wireless multi-point to multi-point links, ADSL or fiber.
Even in the case in which two community networks use the same technology, it is still likely that they use different hardware, firmware or protocols.
The sizes also vary greatly and can range from few tenths to tenths of thousands.
The amount of documentation, supporting applications and organizational structure is also very different in different community networks.

Despite the disparity of models for community networks, they also share some aspects in common.
The most important is the involvement of the users in the planning, financing, deployment and maintainance of the networks.

The aspects that community networks share in common are an interest for connectivity and freedom of choice for the users,

@@

Another area that has witnessed the potential for community-centric innovation is that of networked cryptocurrency.
The combination of cryptography, free libre software and a data network makes it possible the use of networked cryptocurrencies.
In these technologies, a cryptographically secure distributed ledger is maintained by the network.
The network is composed of nodes that interact in a peer-to-peer fashion following a common set of rules or protocol to guarantee the integrity of the cryptocurrency.
The network not only distributes information about the ongoing transactions, it also provides hashing power that is used as a proof of work.
The distributed ledger is protected by the proof of work.
In a sense, the nodes vote with their proof-of-work and the majority decides which transactions are included in the public distributed ledger.

Cryptocurrency has some advantages when compared to traditional fiat currencies.
Some of the advantages of cryptocurrencies are related to their peer-to-peer nature.
One of them is the lack of a central authority.
A central authority can make arbitrary decisions and create money at no cost.
In cryptocurrencies, the emission of money is controlled by the protocol and governed by publicly known rules.
This means that the availability of money at any point in the future is known by all the participants.
The emitted money is used to reward the nodes that contribute with hashing power to the security of the network.
Hashing power requires hardware and electricity, which means that is no longer possible to create money at zero cost.
In fact, the money generation business (popularly known as \emph{mining}) is, in a financial sense, a risky endeavour.
The cost of hashing power required to \emph{mine} can exceed the value of the generated money.

Peer-to-peer networks are extremely resilient.
There is no single point of failure or central entity that can be attacked to take down the network.
As there are thousands of nodes spread across the network, the currency can survive as long as the underlying data network survives.
The risks of blockades or political interference are greatly mitigated.
Many Internet organizations such as Wikileaks and Wordpress accept donations and payments in cryptocurrency.
In some cases to bypass totalitarian governments blockades and in others to operate in countries with lacking payment networks.

Some of the advantages of networked cryptocurrencies are derived from their networked nature.
The reach of the cryptocurrency is the same as that of the underlying network.
If a cryptocurrency runs on the Internet, the scope is global.
As a result, economic transactions benefit from all the advantages of Internet protocols.
Sending money is as simple as sending an electronic mail.
It propagates almost instantaneously across all the Internet and is confirmed in a matter of minutes.
Payments can be easily automatized using scripts or other software.
Any networked device, such as servers, laptops or mobile phones, can be used to send a payment.
Furthermore, the network is available 24 a day 365 days a year.

@@

As cryptocurrencies are based on free libre software, anyone can create a new currency.
If a currency is used by a single individual is not very useful.
The potential of networked cryptocurrencies materializes when they are used by communities.
The communities can create or agree in a currency for their internal exchanges.
As a result, the currency is impregnated by the values and ethical standards of the community.

@@

The gift (or reputation) economy is the kind of economy we find on the Internet, in which the participants offer software, articles, music, pictures and all kind of digital goods for free.
This can be achieved using permissive licenses such as Creative Commons.
In the information age in which is obvious that money is just information and wealth is measured in knowledge instead of material goods, the progressive evolution towards a reputation economy is a natural step.

@@

Throughout human history, organizational models have changed, transformed and evolved.
As technology evolves, old models are been progressively replaced by others that more efficient in the context of the new technologies.
The invention of information processing and transmission tools multiplied the potential of community-centric models.
Initiatives embracing the community centric model have an advantage compared to those attaching to old hierarchical approaches.

It is natural that we find salient examples of community-centric success in the field of information and communication technologies.
The Internet itself is probably the best example of a community-centric endeavour.
There is a large number of people fascinated by and contributing to the Internet.
The decision-making forums are open and anyone is free to contribute expertise.
In a first level the users organize as networks, and then at another level those networks interconnect among them using peering agreements.

Another example of the successful application of the community-centric model is the free libre open source movement.
Thousands of programmers affiliate to and contribute to this movement working as a self-managed community.
The goods produced by this community often outperform those that have been developed using hierarchical models.
By making it possible for anyone to contribute, FLOSS projects have an advantage with respect to those that choose more restrictive models.

The Wikipedia is an Internet-based project that has also adopted the community-centric model.
The users are not mere consumers of information. 
On the contrary, the community of users is the one playing a central role by writing, revising and correcting the encyclopedical articles.
The community-centric model has allowed Wikipedia to grow in terms of number of articles, languages and quality to a point that surpassed encyclopedias that were built using more traditional models.

@@

In the case of open hardware, the fact that all schematics and manufacturing details are publicly available on the Internet makes it very economical for manufacturer to produce the goods.
The manufacturer can avoid all research and development costs and focus only in the fabrication.
As a consequence, open hardware devices can be found at very low prices in the market.
Lower prices translate to a greater community that can contribute to further improve the original product.

@@

The other branch of Commons for Europe is Bottom up Broadband.
This branch is focused on understanding and documenting the working principles of community networks.

\section{Related Work}
\cite{bogliolo2007uwc}
\section{Conclusions}

>>>>

There are cases in which the participants of community networks include not only individuals, but also local organizations such as municipalities or universities \cite{bogliolo2007uwc}.
This involvement stems from the fact that community network have been effective in overcoming the digital divide \cite{oliver2010wca}.

Community networks are particularly useful in underserved areas with little presence of commercial operators.
In rural, distant or poor areas community networks can be the only viable alternative to access the Internet.
In these places, the inhabitants have a strong incentive to spend time and energies in acquiring the necessary skills required to deploy and operate a network.

@@

There is a large number of community networks and they are very heterogeneous.
They use different technologies such as wireless point-to-point links, wireless multi-point to multi-point links, ADSL or fiber.
Even in the case in which two community networks use the same layer-2 technology, it is still likely that they use different hardware, firmware or protocols.
The sizes of community networks also vary greatly and can range from few tenths of nodes to tenths of thousands.
The amount of documentation, supporting applications and organizational structure is also very different in different community networks.

Despite the disparity of models for community networks, there is an aspect that is common to all of them.
All these network are user-centric.
They are planned, financed, deployed and maintained by the same users that benefit from them.

@@

Another area that has witnessed the potential for community-centric innovation is that of networked cryptocurrency.
The combination of cryptography, free libre software and a data network makes it possible the use of networked cryptocurrencies.
In these technologies, a cryptographically secure distributed ledger is maintained by the network.
The network is composed of nodes that interact in a peer-to-peer fashion following a common set of rules (i.e., a protocol) to guarantee the integrity of the cryptocurrency.
The network not only distributes information about the ongoing transactions, it also provides hashing power that is used as a proof of work.
The distributed ledger is protected by the proof of work.
In a sense, the nodes vote with their computational work and the majority decides which transactions are included in the public distributed ledger.

Cryptocurrency has some advantages when compared to traditional fiat currencies.
Some of the advantages of cryptocurrencies are related to their peer-to-peer nature.
One of them is the lack of a central authority.
A central authority can make arbitrary decisions and create money at no cost.
In cryptocurrencies, the emission of money is controlled by the protocol and governed by publicly known rules.
This means that the availability of money at any point in the future is known by all the participants.
The emitted money is used to reward the nodes that contribute with hashing power to the security of the network.
Hashing power requires hardware and electricity, which means that it is no longer possible to create money at zero cost.
In fact, the money generation business (popularly known as \emph{mining}) is, in a financial sense, a risky endeavour.
The cost of hashing power required to \emph{mine} can exceed the value of the generated money.

Peer-to-peer networks are extremely resilient.
There is no single point of failure or central entity that can be attacked to take down the network.
As there are thousands of nodes spread across the network, the currency can survive as long as the underlying data network survives.
The risks of blockades or political interference are greatly mitigated.
Many Internet organizations such as Wikileaks and Wordpress accept donations and payments in cryptocurrency.
In some cases to bypass governments blockades and in others to operate in countries with lacking payment networks.

Some of the advantages of networked cryptocurrencies are derived from their networked nature.
The reach of the cryptocurrency is the same as that of the underlying network.
If a cryptocurrency runs on the Internet, the scope is global.
As a result, economic transactions benefit from all the advantages of Internet protocols.
Sending money is as simple as sending an electronic mail.
A cryptocurrency transaction propagates almost instantaneously across all the Internet and is confirmed in a matter of minutes.
Payments can be easily automatized using scripts or other software.
Any networked device, such as servers, laptops or mobile phones, can be used to send a payment.
Furthermore, the payment network is available 24 hours a day and 365 days a year.

@@

As cryptocurrencies are based on free libre software, anyone can create a new currency.
If a currency is used by a single individual it is not very useful.
The potential of networked cryptocurrencies materializes when they are used by communities.
The communities can create or agree in a currency for their internal exchanges.
As a result, the currency is impregnated by the values and ethical standards of the community.
In a sense, the value of cryptocurrencies is backed only by the community of users.

@@

The gift (or reputation) economy is the kind of economy we find on the Internet, in which the participants offer software, articles, music, pictures and all kinds of digital goods for free.
This can be achieved using permissive licenses such as Creative Commons.
In the information age money is just information and wealth is measured in knowledge instead of material goods.
The progressive evolution towards a reputation economy is a natural step.

@@

Throughout human history, organizational models have changed, transformed and evolved.
As new technologies develop, old models are progressively replaced by others that are more efficient in the context of the new technologies.
The invention of information processing and transmission tools multiplied the potential of community-centric models.
Initiatives embracing the community-centric model have an advantage compared to those attaching to old hierarchical approaches.

It is natural that we find salient examples of community-centric success in the field of information and communication technologies.
The Internet itself is probably the best example of a community-centric endeavour.
There is a large number of people fascinated by and contributing to the Internet.
The decision-making forums are open and anyone is free to contribute expertise.

Another example of the successful application of the community-centric model is the free libre open source movement.
Thousands of programmers affiliate to and contribute to this movement working as a self-managed community.
The software produced by this community often outperforms what have been developed using hierarchical models.
By making it possible for anyone to contribute, FLOSS projects have an advantage with respect to those that choose more restrictive models.

The Wikipedia is an Internet-based project that has also adopted the community-centric model.
The users are not mere consumers of information. 
On the contrary, the community of users is the one playing a central role by writing, revising and correcting the encyclopedical articles.
The community-centric model has allowed Wikipedia to grow in terms of number of articles, languages and quality.
In certain aspects, the community-centric encyclopedia has surpassed encyclopedias that were built using more traditional models.

@@

In the case of open hardware, the fact that all schematics and manufacturing details are publicly available on the Internet makes it very economical for a manufacturer to produce the goods.
The manufacturer can avoid all research and development costs and focus only in the fabrication.
As a consequence, open hardware devices can be found at very low prices in the market.
Lower prices translate to a greater community that can contribute to further improve the original product.

@@

The other branch of the Commons for Europe project is called Bottom up Broadband.
This branch is focused on understanding and documenting the working principles of community networks.

\subsection{Confine}
The Confine project emphasizes the links between research and community networks.
The goal is to make it easier for researchers to run experiments in community networks.
It provides the tools for automatize and monitor the experiments, as well as the hardware distributed across the community network.

Community networks have grown to a size far beyond what can be reproduced in a lab.
Therefore, the most effective way to learn and understand this phenomena is by carrying out research directly on the community network.
It is a good opportunity for researchers to be in close contact with these networks that are leading the innovation in some areas of networking.
At the same time, the community network benefits from the expertise of the research community and from the documentation and dissemination efforts associated to the research process.

\section{Related Work}
\section{Conclusions}
